---
title: "Working with a data frame"
format: 
  revealjs:
    code-fold: false
    theme: simple
    footer: "GEO4251 Climate Change & Storms"
    chalkboard: true
editor: source
---

## 

* Throughout the semester you'll focus on data that are saved in a “spreadsheet”-type format within your environment called a _data frame_

* Inside a _data frame_ rows are observations and columns are variables

* You start be considering a data frame listing hurricane counts by year in the United States

* <http://myweb.fsu.edu/jelsner/US.txt>

* The data are updated annually. For example, last year both Hurricane Ian and Hurricane Nicole hit Florida

##

```{r}
#| echo: true
L <- "http://myweb.fsu.edu/jelsner/US.txt"
LH.df <- readr::read_table(file = L)

LH.df |>
  head()
```

`Year`, `All`, `MUS`, `G`, `FL` and `E` are different columns in the data frame. These are the variables. Each row is an observation over an entire year.

##

Using the `View()` function brings up RStudio's built-in data viewer
```{r, eval=FALSE}
#| echo: true
LH.df |>
  View()
```

## {.smaller}

The `dplyr::glimpse()` function is used to view all the columns
```{r}
#| echo: true
LH.df |>
  dplyr::glimpse()
```

## {.smaller}

Recall the tornado data. Note the different variable types `<dbl>`, `<chr>`, ...
```{r}
#| echo: true
"http://www.spc.noaa.gov/wcm/data/1950-2021_actual_tornadoes.csv" |>
  readr::read_csv(show_col_types = FALSE) |>
  dplyr::glimpse()
```

## {.smaller}

The `$` extraction operator is used to view single variables/columns

```{r}
#| echo: true
LH.df$FL
```
