---
title: "Lab 1"
author: "ANSWERS"
format: pdf
editor: source
---

## 1 
Write code below to download and install the {ggplot2} package
```{r, eval=FALSE}
#| echo: true
install.packages("ggplot2")
library(ggplot2)
```

## 2 
Write code below to get the tornado data into your environment as a data frame called `Torn.df`
```{r}
#| echo: true
L <- "http://www.spc.noaa.gov/wcm/data/1950-2021_actual_tornadoes.csv"
Torn.df <- readr::read_csv(file = L,
                           show_col_types = FALSE)
```

## 3 
Write code below to get the average tornado path length. The path length variable is `len` in `Torn.df`
```{r}
#| echo: true
Torn.df$len |>
  mean()
```

## 4
Write code below to get the number of tornadoes in each state. Hint: use the `table()` function on the variable `st`
```{r}
#| echo: true
Torn.df$st |>
  table()
```

## 5

Type your answers below to the following questions: According to the PBS video entitled "What is the riskiest region in the US as the climate changes?" (linked in the file `05-working-with-data.qmd`), what state has the county with the highest overall risk resulting from climate change? What is the name of the county? (2)

Beaufort County, South Carolina