---
title: "Quantifying changes over time"
format: 
  revealjs:
    code-fold: false
    theme: simple
    footer: "GEO4251 Climate Change & Storms"
    chalkboard: true
editor: source
---

##

Returning to the rain data from Northwest Florida 

-   Add a column with June rain values in units of centimeters
-   Make a time series plot showing the values over the years
-   Include a trend line

```{r}
#| echo: true
#| output-location: slide
Rain.df <- readr::read_table(here::here("data", "NWFL_MonthlyRain.txt")) |>
  dplyr::select(Year, Rain_in = Jun) |>
  dplyr::mutate(Rain_cm = Rain_in * 2.54)

library(ggplot2)

ggplot(data = Rain.df,
       mapping = aes(x = Year, y = Rain_cm)) +
  geom_line() + scale_x_continuous(breaks = seq(1900, 2020, 20)) +
  geom_smooth(method = lm, se = FALSE) +
  ylab("Rain (cm)")
```

##

-   The slope of the blue line is the average amount of change over the years
-   You quantify this change using the following code

```{r}
#| echo: true
lm(Rain_cm ~ Year, data = Rain.df)
```
-   The value below `Year` is the increase in the amount of rain in centimeters per year
-   +.023 cm per year or +2.3 cm per 100 years

## A key climate indicator

-   Temperature is a key climate indicator. Temperature measurements made on land and in water across the globe tell us how temperature has changed over years, decades and centuries. They also tell us how temperature varies from one place to another

-   Global temperature is an aggregate over all the local temperature changes. It is a summary of the overall tendency in surface temperature and it is probably the longest instrumental series of global climate change that we have

##

-   Global temperatures from various research labs <https://climate.metoffice.cloud/temperature.html#datasets>

-   Annual global mean temperatures expressed as a difference from pre-industrial conditions (anomalies). Four data sets are available and there is good agreement between them

-   Anomalies are calculated relative to a 1981 to 2010 baseline and offset by 0.69°C, which is the best estimate difference for that period from the 1850-1900 average given in the IPCC sixth assessment report

## 

Global mean temperature

```{r}
#| echo: true
L <- "https://climate.metoffice.cloud/formatted_data/gmt_HadCRUT5.csv"
download.file(url = L,
              destfile = here::here("data", "T.csv"))

T.df <- readr::read_csv(here::here("data", "T.csv"),
                        show_col_types = FALSE)
names(T.df) <- c("Year", "T_C", "u")

T.df |>
  head()
```

## {.smaller}

-   Create a time series graph of the temperature anomalies. Include a ribbon indicating the uncertainty around the estimated values

```{r}
#| echo: true
ggplot(data = T.df,
       mapping = aes(x = Year, y = T_C, ymin = T_C-u, ymax = T_C+u)) +
  geom_ribbon(fill = "gray") + 
  geom_line() +
  scale_x_continuous(breaks = seq(1860, 2020, 20)) +
  ylab("°C")
```

##

-   Overall, average temperatures have increased in most areas since the start of the 20th Century. This is reflected in the long-term increase in global temperature
-   However, temperatures have not increased at the same rate everywhere
-   Some general patterns of warming are to be noted: the land has generally warmed faster than the oceans; and over the past two decades, the Arctic has warmed twice as fast as the rest of the world


##

-   The increase in global temperatures is not steady
-   Slight cooling until about 1920 followed by a warming until the mid 1940s
-   No warming from mid 1940s until the late 1970s
-   Nearly linear warming since then
-   These variations in the rate of warming/cooling can be illustrated using a local smoother

##

```{r}
#| echo: true
#| code-line-numbers: "4"
ggplot(data = T.df,
       mapping = aes(x = Year, y = T_C)) +
  geom_line() +
  geom_smooth(span = .5, se = FALSE) +
  scale_x_continuous(breaks = seq(1860, 2020, 20)) +
  ylab("°C")
```


##

Let's re-plot the annual temperatures starting in 1973 and then add a trend line

```{r}
#| echo: true
T.df |>
  dplyr::filter(Year >= 1973) |>
ggplot(mapping = aes(x = Year, y = T_C)) +
  geom_line() +
  geom_smooth(method = lm, se = FALSE) +
  scale_x_continuous(breaks = seq(1980, 2020, 10)) +
  ylab("°C")
```

##

-   The slope of the blue line is the average increase in temperatures over the years
-   You quantify this change using the following code

```{r}
#| echo: true
lm(T_C ~ Year, data = T.df[T.df$Year >= 1973, ])
```
-   The value below `Year` is the increase in temperature in °C per year
-   +.019°C per year or about +1.9°C per 100 years

##

Global humidity data <https://climate.metoffice.cloud/humidity.html#datasets>

Specific humidity (q in g/kg)
```{r}
#| echo: true
L <- "https://climate.metoffice.cloud/formatted_data/humidity_q_HadISDH%20blend%20q.csv"
download.file(url = L,
              destfile = here::here("data", "q.csv"))

q.df <- readr::read_csv(here::here("data", "q.csv"),
                        show_col_types = FALSE)
names(q.df) <- c("Year", "q_gpkg", "u")

q.df |>
  head()
```

##

```{r}
#| echo: true
q.df |>
ggplot(mapping = aes(x = Year, y = q_gpkg)) +
  geom_line() +
  geom_smooth(method = lm, se = FALSE) +
  scale_x_continuous(breaks = seq(1980, 2020, 10)) +
  ylab("g/kg")
```

##

-   The slope of the blue line is the average increase in specific humidity over the years
-   You quantify this change using the following code

```{r}
#| echo: true
lm(q_gpkg ~ Year, data = q.df)
```

-   The value below `Year` is the increase in specific humidity in g/kg per year
-   +.008 g/kg per year


##

Relative humidity

```{r}
#| echo: TRUE
download.file(url = "https://climate.metoffice.cloud/formatted_data/humidity_RH_HadISDH%20blend%20RH.csv",
              destfile = here::here("data", "RH.csv"))

RH.df <- readr::read_csv(here::here("data", "RH.csv"),
                        show_col_types = FALSE)
names(RH.df) <- c("Year", "RH_p", "u")

RH.df |>
  head()
```

##

```{r}
#| echo: true
RH.df |>
ggplot(mapping = aes(x = Year, y = RH_p)) +
  geom_line() +
  geom_smooth(method = lm, se = FALSE) +
  scale_x_continuous(breaks = seq(1980, 2020, 10)) +
  ylab("%")
```
##

-   The slope of the blue line is the average decrease in relative humidity over the years
-   You quantify this change using the following code

```{r}
#| echo: true
lm(RH_p ~ Year, data = RH.df)
```

-   The value below `Year` is the decrease in relative humidity in percentage points per year
-   -1.3 percentage points per 100 years
-   Why the decrease?

##

Dew point temperature

```{r}
#| echo: TRUE
download.file(url = "https://climate.metoffice.cloud/formatted_data/humidity_Td_HadISDH%20blend%20Td.csv",
              destfile = here::here("data", "Td.csv"))

Td.df <- readr::read_csv(here::here("data", "Td.csv"),
                        show_col_types = FALSE)
names(Td.df) <- c("Year", "Td_C", "u")

Td.df |>
  head()
```

##

```{r}
#| echo: true
Td.df |>
ggplot(mapping = aes(x = Year, y = Td_C)) +
  geom_line() +
  geom_smooth(method = lm, se = FALSE) +
  scale_x_continuous(breaks = seq(1980, 2020, 10)) +
  ylab("°C")
```

##

-   The slope of the blue line is the average increase in dew point temperature over the years
-   You quantify this change using the following code

```{r}
#| echo: true
lm(Td_C ~ Year, data = Td.df)
```

-   The value below `Year` is the increase in dew point in °C per year
-   +1.3°C per 100 years

##

Explore the relationships between humidity and temperature

```{r}
#| echo: true
dplyr::left_join(q.df, T.df, by = "Year") |>
  ggplot(mapping = aes(x = T_C, y = q_gpkg)) +
  geom_point() +
  geom_smooth(method = lm)
```
