---
title: "Weather and climate data"
format: 
  revealjs:
    code-fold: false
    theme: white
    footer: "GEO4251 Climate Change & Storms"
    chalkboard: true
editor: source
---

## 

-   Reliable resources for weather and climate data
-   Quantify changes in extremes
-   Humidity
-   Early storm chase video

## Reliable resources for weather and climate data {.smaller}

-   [Today's weather](https://climatereanalyzer.org/wx/todays-weather/)
-   [Weather models](https://www.tropicaltidbits.com/analysis/models/)
-   [Climate change indicators](https://zacklabe.com/climate-change-indicators/)
-   [Tracking the changing climate with Earth observations](https://climate.metoffice.cloud/)
-   [International Best Track Archive for Climate Stewardship (IBTrACS)](https://www.ncei.noaa.gov/products/international-best-track-archive)
-   [NOAA NWS Storm Prediction Center](https://www.spc.noaa.gov/)
-   [Global reanalysis data](https://cds.climate.copernicus.eu/cdsapp#!/dataset/reanalysis-era5-single-levels-monthly-means?tab=overview)

##

-   The data science revolution is the increasing importance and widespread adoption of data-driven decision-making
-   It is a result of massive amounts of data generated and collected by businesses, organizations, and individuals
-   And advancements in computing power, storage, and algorithms enabling the processing and analysis of these data
-   Data science is the use of techniques to extract insights and knowledge from data, and to inform and optimize decision-making processes
-   It draws on statistics, computer science, mathematics, and domain-specific knowledge, and makes use of machine learning, data mining, and visualization

##

-   It has transformed finance, healthcare, retail, and marketing 
-   It has enabled businesses to gain a deeper understanding of their customers, optimize their operations, and develop new products and services
-   It has opened up new opportunities for research and innovation in areas from climate science to social science
-   Before the revolution it was seen as inappropriate for people “in the business” to write code
-   Code was a black art reserved for specially trained wizards in information technology development teams
-   Now in most organizations it’s normal to run R or Python in the business

## Quantify changes in extremes

-   Rain data from NW Florida

```{r warning=FALSE, message=FALSE}
#| echo: true
Rain.df <- readr::read_table(here::here("data", "NWFL_MonthlyRain.txt")) |>
  dplyr::select(Year, Jun) |>
  dplyr::mutate(Rain_cm = Jun * 2.54)
```

##

-   Trend line showing changes to the average monthly rain amount

```{r warning=FALSE, message=FALSE}
#| echo: true
library(ggplot2)

ggplot(data = Rain.df,
       mapping = aes(x = Year, y = Rain_cm)) +
  geom_line() +
  geom_smooth(method = lm)
```
##

-   Increasing extremes

```{r warning=FALSE, message=FALSE}
#| echo: true
ggplot(data = Rain.df,
       mapping = aes(x = Year, y = Rain_cm)) +
  geom_line() +
  geom_quantile(quantiles = c(.25, .5, .75, .95))
```

-   Air can hold an additional 7% more water vapor for every one degree C increase in temperature

## Humidity

-   Water is special
-   Water exists in all three phases on Earth
-   Relative vs absolute humidity

{{< video https://www.youtube.com/embed/wB3Yyq4xy_Y >}} 3 minutes

## {.smaller}

-   Relative humidity and dew point temperature are measures of the amount of moisture in the air, but they represent different aspects of that moisture
-   Relative humidity is a measure of how much moisture is in the air relative to the maximum amount it could hold at that temperature
    -   Expressed as a percentage indicating how close the air is to being saturated with moisture
    -   If the relative humidity is 50%, the air is holding half of the maximum amount of moisture it could hold at that temperature
    
##

-   Dew point temperature, on the other hand, is the temperature at which the air would become saturated with moisture, causing dew to form
    -   It is a measure of the actual amount of moisture in the air, regardless of the temperature
-   When the dew point temperature and the air temperature are close, the air is close to being saturated and the relative humidity is high
-   When the dew point temperature and air temperature are far apart, the air is relatively dry and the relative humidity is low

## 

-   Global humidity data <https://climate.metoffice.cloud/humidity.html#datasets>
-   Relative humidity data 

```{r}
#| echo: TRUE
download.file(url = "https://climate.metoffice.cloud/formatted_data/humidity_RH_HadISDH%20blend%20RH.csv",
              destfile = here::here("data", "RH.csv"))

RH.df <- readr::read_csv(here::here("data", "RH.csv"),
                        show_col_types = FALSE)
names(RH.df) <- c("Year", "RH_p", "u")

RH.df |>
  head()
```

##

```{r}
#| echo: true

RH.df |>
ggplot(mapping = aes(x = Year, y = RH_p)) +
  geom_line() +
  geom_smooth(method = lm, se = FALSE) +
  scale_x_continuous(breaks = seq(1980, 2020, 10)) +
  ylab("%")
```

## 

-   The slope of the blue line is the average decrease in relative humidity over the years
-   You quantify this change using the following code

```{r}
#| echo: true
lm(RH_p ~ Year, data = RH.df)
```

-   The value below `Year` is the decrease in relative humidity in percentage points per year
-   -1.3 percentage points per 100 years
-   Why the decrease?

## 

-   Dew point temperature data

```{r}
#| echo: TRUE
download.file(url = "https://climate.metoffice.cloud/formatted_data/humidity_Td_HadISDH%20blend%20Td.csv",
              destfile = here::here("data", "Td.csv"))

Td.df <- readr::read_csv(here::here("data", "Td.csv"),
                        show_col_types = FALSE)
names(Td.df) <- c("Year", "Td_C", "u")

Td.df |>
  head()
```

## 

```{r}
#| echo: true
Td.df |>
ggplot(mapping = aes(x = Year, y = Td_C)) +
  geom_line() +
  geom_smooth(method = lm, se = FALSE) +
  scale_x_continuous(breaks = seq(1980, 2020, 10)) +
  ylab("°C")
```

## 

-   The slope of the blue line is the average increase in dew point temperature over the years
-   You quantify this change using the following code

```{r}
#| echo: true
lm(Td_C ~ Year, data = Td.df)
```

-   The value below `Year` is the increase in dew point in °C per year
-   +1.3°C per 100 years

## 

-   Specific humidity (g/kg) data

```{r}
#| echo: true
L <- "https://climate.metoffice.cloud/formatted_data/humidity_q_HadISDH%20blend%20q.csv"
download.file(url = L,
              destfile = here::here("data", "q.csv"))

q.df <- readr::read_csv(here::here("data", "q.csv"),
                        show_col_types = FALSE)
names(q.df) <- c("Year", "q_gpkg", "u")

q.df |>
  head()
```

## 

```{r}
#| echo: true

q.df |>
ggplot(mapping = aes(x = Year, y = q_gpkg)) +
  geom_line() +
  geom_smooth(method = lm, se = FALSE) +
  scale_x_continuous(breaks = seq(1980, 2020, 10)) +
  ylab("g/kg")
```

## 

-   The slope of the blue line is the average increase in specific humidity over the years
-   You quantify this change using the following code

```{r}
#| echo: true
lm(q_gpkg ~ Year, data = q.df)
```

-   The value below `Year` is the increase in specific humidity in g/kg per year
-   +.008 g/kg per year

## 

-   The relationship between humidity and temperature
-   Get the temperature data

```{r}
#| echo: true
L <- "https://climate.metoffice.cloud/formatted_data/gmt_HadCRUT5.csv"
download.file(url = L,
              destfile = here::here("data", "T.csv"))

T.df <- readr::read_csv(here::here("data", "T.csv"),
                        show_col_types = FALSE) 
names(T.df) <- c("Year", "T_C", "u")
```

## 

-   Join the temperature data to the humidity data

```{r}
#| echo: true
dplyr::left_join(q.df, T.df, by = "Year") |>
  ggplot(mapping = aes(x = T_C, y = q_gpkg)) +
  geom_point() +
  geom_smooth(method = lm)
```

##

-   Storm chasing with my kids in 2008

<https://www.dropbox.com/s/qdr9849tjpn6mv6/May23_2008_StormChase2.mov?dl=0>

## 

-   A tornado season forecast for 2023

{{< video https://www.youtube.com/embed/5kVeEf7CtCE >}} 8 minutes
