---
title: "Increasing temperature, rising seas, melting ice"
format: 
  revealjs:
    slide-number: c/t
    code-fold: false
    chalkboard: true
    theme: white
    footer: "GEO4251 Climate Change & Storms"
editor: source
---

## 

-   Temperature changes across time
-   The climate stripes and hockey stick graphs
-   Rising seas and melting ice
-   Local temperature changes
-   Lapse rates of temperature
-   Today's winds <https://weather.rap.ucar.edu/surface/>

##  {.smaller}

-   Get the global average temperature data by month using the `get_temp()` function from the {hockeystick} package

```{r}
#| echo: true
library(hockeystick)

t.df <- get_temp()

t.df
```

## 

-   Make a time-series plot

```{r}
#| echo: true
#| fig-cap: "Annual global land-ocean temperature index"
t.df |>
  plot_temp()
```

## 

-   Plot the time-series as stripes (warming stripes)

```{r}
#| echo: true
t.df |>
  warming_stripes()
```

-   International symbol of global warming (Dr. Ed Hawkins) {{< video https://www.youtube.com/embed/3VIHRLUnVOI >}} 3 minutes

##  {.smaller}

-   Convert the wide data frame to a long data frame

```{r}
#| echo: true
t.dfL <- t.df |>
  dplyr::select(Year, month.abb) |>
  tidyr::pivot_longer(cols = month.abb,
                      names_to = "Month",
                      values_to = "Anom_C") |>
  dplyr::mutate(Year = lubridate::year(Year),
                Month = factor(Month, 
                               levels = month.abb))
t.dfL
```

## 

-   Plot temperature trends by month since 1980

```{r}
#| echo: true
library(ggplot2)

t.dfL |>
  na.omit() |>
  dplyr::filter(Year >= 1980) |>
  ggplot(mapping = aes(x = Year, 
                       y = Anom_C)) +
  geom_line() +
  geom_smooth(method = lm, color = "red") +
  facet_wrap(~ Month, ncol = 4)
```

## 

-   Get a temperature index since the birth of Christ (Common Era)
-   Combined modeled, instrumental, and proxy data

```{r}
#| echo: true
anomaly2k <- get_temp2k()

anomaly2k 
```

## 

```{r}
#| echo: true
anomaly2k |>
  plot_temp2k()
```

## 

-   Warming and emissions (Dr. Jennifer Francis) {{< video https://www.youtube.com/embed/JKrkiZM1asQ >}} 5 minutes
-   The hockey stick graph (Simon Clark) {{< video https://www.youtube.com/embed/CqtZdnpfgIc >}} 9 minutes

## 

-   Get the sea level data using the `get_sealevel()` function from the {hockeystick} package

```{r}
#| echo: true
gmsl <- get_sealevel()

gmsl
```

## 

-   Make a time-series plot

```{r}
#| echo: true
gmsl |>
  plot_sealevel()
```

##  {.smaller}

-   Get the sea ice data using the `get_seaice()` function from the {hockeystick} package
-   Sea ice extent in millions of square kilometers every July 31 since 1979

```{r}
#| echo: true
seaice <- get_seaice() 

seaice 
```

## 

-   Make a time-series plot

```{r}
#| echo: true
seaice |>
  plot_seaice()
```

## 

-   Quantify the loss of sea ice in square kilometers per year

```{r}
#| echo: true
seaice <- seaice |>
  dplyr::mutate(Year = lubridate::year(date),
                Extent_km2 = extent * 10^6)

lm(Extent_km2 ~ Year, data = seaice)
```

-   A loss of more than 68 thousand square kilometers per year

##

-   Paleo ice core

```{r}
#| echo: true
vostok <- get_paleo()

vostok |>
  plot_paleo()
```

## 

-   All together

```{r}
#| echo: true
climate_grid()
```

##  {.smaller}

-   Regional warming
-   Is Tallahassee getting warmer?
-   Get the daily high temperature from the official airport site since 1940

```{r, message=FALSE}
#| echo: true
L <- "http://myweb.fsu.edu/jelsner/temp/data/TLH_Daily1940-2020.csv"
TLH_Daily.df <- readr::read_csv(L) |>
  dplyr::mutate(Year = lubridate::year(DATE)) |>
  dplyr::filter(Year >= 1969 & Year < 2021)

TLH_Daily.df
```

##  {.smaller}

-   Graph the annual average temperature

```{r, message=FALSE}
#| echo: true
TLH_Annual.df <- TLH_Daily.df |>
  dplyr::group_by(Year) |>
  dplyr::summarize(AvgT = mean(TMAX, na.rm = TRUE))

ggplot(data = TLH_Annual.df,
       mapping = aes(x = Year, y = AvgT)) +
  geom_point() +
  geom_line() +
  geom_smooth(method = lm)
```

## 

-   Quantify the change

```{r}
#| echo: true
lm(AvgT ~ Year, data = TLH_Annual.df)
```

##  {.smaller}

-   Is Las Vegas getting warmer?
-   Get the daily high temperature from the official airport site since 1969

```{r, message=FALSE}
#| echo: true
L <- "http://myweb.fsu.edu/jelsner/temp/data/LVG_Daily1940-2020.csv"
LVG_Daily.df <- readr::read_csv(L) |>
  dplyr::mutate(DATE = readr::parse_datetime(DATE, "%m/%d/%y")) |>
  dplyr::mutate(Year = lubridate::year(DATE)) |>
  dplyr::filter(Year >= 1969 & Year < 2021)

LVG_Daily.df
```

##  {.smaller}

-   Graph the annual average temperature

```{r, message=FALSE}
#| echo: true
LVG_Annual.df <- LVG_Daily.df |>
  dplyr::group_by(Year) |>
  dplyr::summarize(AvgT = mean(TMAX, na.rm = TRUE))

ggplot(data = LVG_Annual.df,
       mapping = aes(x = Year, y = AvgT)) +
    geom_point() +
    geom_line() +
    geom_smooth(method = lm)
```

## 

-   Quantify the change

```{r}
#| echo: true
lm(AvgT ~ Year, data = LVG_Annual.df)
```

##  {.smaller}

-   Is the frequency of extremely hot days increasing?
-   Here you define a hot day as one in which the high temperatures reaches 100째F

```{r, message=FALSE}
#| echo: true
library(MASS)

TLH_Daily.df |>
  dplyr::group_by(Year) |>
  dplyr::summarize(N100 = sum(TMAX >= 100, na.rm = TRUE)) |>
ggplot(mapping = aes(x = Year, y = N100)) +
  geom_col() +
  geom_smooth(method = "glm.nb", col = "orange") +
  ylab("Number of days in Tallahassee at or exceeding 100째F")
```

##  {.smaller}

-   Compare the frequency of daily high temperatures between Tallahassee and Las Vegas

```{r, message=FALSE}
#| echo: true
( TLH.p <- TLH_Daily.df |>
  dplyr::group_by(TMAX) |>
  dplyr::summarize(nH = dplyr::n()) |>
ggplot(mapping = aes(x = TMAX, y = nH)) +
  geom_col(col = 'white', fill = "gray70") +
  labs(title = "Frequency of Daily High Temperatures",
       subtitle = "Tallahassee, FL, USA (1969-2020)",
       x = "Daily High Temperature (째F)",
       y = "Number of Days") +
 scale_x_continuous(breaks = seq(from = 20, to = 110, by = 10)) +
 theme_minimal() )
```

##  {.smaller}

```{r, message=FALSE}
#| echo: true
( LVG.p <- LVG_Daily.df |>
  dplyr::group_by(TMAX) |>
  dplyr::summarize(nH = dplyr::n()) |>
ggplot(mapping = aes(x = TMAX, y = nH)) +
  geom_col(col = 'white', fill = "gray70") +
  labs(title = "Frequency of Daily High Temperatures",
       subtitle = "Las Vegas, NV, USA (1969-2020)",
       x = "Daily High Temperature (째F)",
       y = "Number of Days") +
 scale_x_continuous(breaks = seq(from = 20, to = 110, by = 10)) +
 theme_minimal() )
```

## 

-   Plot together

```{r}
#| echo: true
library(patchwork)

TLH.p / LVG.p
```

-   Why the difference?

## 

-   Lapse rate of temperature
-   Parcel theory (dry vs moist lapse rates)
