---
title: "More CO2, increasing temperature, rising seas, melting ice"
format: 
  revealjs:
    slide-number: c/t
    code-fold: false
    chalkboard: true
    theme: white
    footer: "GEO4251 Climate Change & Storms"
editor: source
---

##

-   More CO~2~
-   Increasing temperature
-   Rising seas
-   Melting ice

## 

-   Download the temperature data using the `get_temp()` function from the {hockeystick} package

```{r}
#| echo: true
library(hockeystick)

t.df <- hockeystick::get_temp(write_cache = TRUE)

t.df
```

## 

-   Make a time-series plot

```{r}
#| echo: true
#| fig-cap: "Annual global land-ocean temperature index"
t.df |>
  plot_temp()
```

##

-   Plot the time-series as stripes (warming stripes)

```{r}
t.df |>
  warming_stripes()
```

## 

-   Convert the wide data to long data

```{r}
#| echo: true
t.dfL <- t.df |>
  dplyr::select(Year, month.abb) |>
  tidyr::pivot_longer(cols = month.abb,
                      names_to = "Month",
                      values_to = "Anom_C") |>
  dplyr::mutate(Year = lubridate::year(Year),
                Month = factor(Month, 
                               levels = month.abb))
t.dfL
```

## 

-   Plot temperatures by month

```{r}
#| echo: true
library(ggplot2)

t.dfL |>
  na.omit() |>
  dplyr::filter(Year >= 1980) |>
  ggplot(mapping = aes(x = Year, 
                       y = Anom_C)) +
#  geom_line() +
  geom_smooth(method = lm, color = "red") +
  facet_wrap(~ Month, ncol = 4)
```

## 

-   Common era temperatures

```{r}
#| echo: true
anomaly2k <- get_temp2k()

anomaly2k 
```

##

```{r}
#| echo: true
anomaly2k |>
  plot_temp2k()
```

##

-   Dr. Jennifer Francis on warming and emissions

{{< video https://www.youtube.com/embed/JKrkiZM1asQ >}} 5 minutes

## 

-   Download the sea level data using the `get_sealevel()` function from the {hockeystick} package

```{r}
#| echo: true
gmsl <- get_sealevel(write_cache = TRUE)

gmsl
```

## 

-   Make a time-series plot

```{r}
#| echo: true
gmsl |>
  plot_sealevel()
```

##

-   Download the sea ice data using the `get_seaice()` function from the {hockeystick} package
-   Sea ice extent in millions of square kilometers every July 31 since 1979

```{r}
#| echo: true
seaice <- get_seaice(write_cache = TRUE)

seaice 
```

## 

-   Make a time-series plot

```{r}
#| echo: true
seaice |>
  plot_seaice()
```

## 

-   Quantify the change

```{r}
#| echo: true
seaice <- seaice |>
  dplyr::mutate(Year = lubridate::year(date),
                Extent = extent * 10^6)

lm(Extent ~ Year, data = seaice)
```

## 

-   Download the Arctic ice data using the `get_icecurves()` function from the {hockeystick} package
-   Arctic ice extent in millions of square kilometers every month since 1979.

```{r}
#| echo: true
arcticice <- get_icecurves()
arcticice 
```

## 

-   Make a time-series plot

```{r}
#| echo: true
arcticice |>
  plot_icecurves()
```

## 

-   Paleo ice core

```{r}
#| echo: true
vostok <- get_paleo()

vostok |>
  plot_paleo()
```

##

-   All together

```{r}
#| echo: true
climate_grid()
```

##

-   Is Tallahassee getting hotter?
-   Get the daily high temperature from the official airport site since 1940
```{r, message=FALSE}
#| echo: true
L <- "http://myweb.fsu.edu/jelsner/temp/data/TLH_Daily1940-2020.csv"
TLH_Daily.df <- readr::read_csv(L) |>
  dplyr::mutate(Year = lubridate::year(DATE)) |>
  dplyr::filter(Year >= 1969 & Year < 2021)

TLH_Daily.df
```

##

-   Graph the annual average temperature
```{r, message=FALSE}
#| echo: true
library(ggplot2)

TLH_Annual.df <- TLH_Daily.df |>
  dplyr::group_by(Year) |>
  dplyr::summarize(AvgT = mean(TMAX, na.rm = TRUE))

ggplot(data = TLH_Annual.df,
       mapping = aes(x = Year, y = AvgT)) +
  geom_point() +
  geom_line() +
  geom_smooth(method = lm)
```

```{r}
lm(AvgT ~ Year, data = TLH_Annual.df)
```

##

-   Is Las Vegas getting hotter?
-   Get the daily high temperature from the official airport site since 1969
```{r, message=FALSE}
#| echo: true
L <- "http://myweb.fsu.edu/jelsner/temp/data/LVG_Daily1940-2020.csv"
LVG_Daily.df <- readr::read_csv(L) |>
  dplyr::mutate(DATE = readr::parse_datetime(DATE, "%m/%d/%y")) |>
  dplyr::mutate(Year = lubridate::year(DATE)) |>
  dplyr::filter(Year >= 1969 & Year < 2021)

LVG_Daily.df
```

-   Graph the annual average temperature
```{r, message=FALSE}
#| echo: true
LVG_Annual.df <- LVG_Daily.df |>
  dplyr::group_by(Year) |>
  dplyr::summarize(AvgT = mean(TMAX, na.rm = TRUE))

ggplot(data = LVG_Annual.df,
       mapping = aes(x = Year, y = AvgT)) +
    geom_point() +
    geom_line() +
    geom_smooth(method = lm)
```

```{r}
lm(AvgT ~ Year, data = LVG_Annual.df)
```

##

Is the frequency of extremely hot days increasing? Here we define a hot day as one in which the high temperatures reaches 100째 F

```{r}
TLH_Daily.df |>
  dplyr::group_by(Year) |>
  dplyr::summarize(N100 = sum(TMAX >= 100)) |>
ggplot(mapping = aes(x = Year, y = N100)) +
  geom_col() +
  ylab("Number of days in Tallahassee at or exceeding 100째 F")
```

The hots are getting hotter

##

```{r, message=FALSE}
TLH_Daily.df |>
  dplyr::group_by(TMAX) |>
  dplyr::summarize(nH = dplyr::n()) |>
ggplot(mapping = aes(x = TMAX, y = nH)) +
  geom_col(col = 'white', fill = "gray70") +
  labs(title = "Frequency of Daily High Temperatures",
       subtitle = "Tallahassee, FL, USA (1969-2020)",
       x = "Daily High Temperature (째F)",
       y = "Number of Days") +
 scale_x_continuous(breaks = seq(from = 20, to = 110, by = 10)) +
 theme_minimal()
```

##

```{r, message=FALSE}
LVG_Daily.df |>
  dplyr::group_by(TMAX) |>
  dplyr::summarize(nH = dplyr::n()) |>
ggplot(mapping = aes(x = TMAX, y = nH)) +
  geom_col(col = 'white', fill = "gray70") +
  labs(title = "Frequency of Daily High Temperatures",
       subtitle = "Las Vegas, NV, USA (1969-2020)",
       x = "Daily High Temperature (째F)",
       y = "Number of Days") +
 scale_x_continuous(breaks = seq(from = 20, to = 110, by = 10)) +
 theme_minimal()
```