---
title: "Rising seas and melting ice"
format: 
  revealjs:
    slide-number: c/t
    code-fold: false
    chalkboard: true
    theme: white
    footer: "GEO4251 Climate Change & Storms"
editor: source
---


## Get the sea level data

Download the sea level data using the `get_sealevel()` function from the {hockeystick} package.

```{r}
#| echo: true
library(hockeystick)

gmsl <- get_sealevel(write_cache = TRUE)
```

##

Look at the first & last six rows of using the `head()` function.

```{r}
#| echo: true
gmsl |>
  head()

gmsl |>
  tail()
```

## Make a time-series plot

```{r}
#| echo: true
gmsl |>
  plot_sealevel()
```

## Get the sea ice data

Download the sea ice data using the `get_seaice()` function from the {hockeystick} package.

```{r}
#| echo: true
library(hockeystick)

seaice <- get_seaice(write_cache = TRUE)
```

##

Look at the data using the `head()` function.

```{r}
#| echo: true
seaice |>
  head()
 
```
Sea ice extent in millions of square kilometers every July 31 since 1979.

## Make a time-series plot

```{r}
#| echo: true
seaice |>
  plot_seaice()
```

## Quantify the change

```{r}
#| echo: true
seaice <- seaice |>
  dplyr::mutate(Year = lubridate::year(date),
                Extent = extent * 10^6)

lm(Extent ~ Year, data = seaice)
```

## Get the Arctic ice data

Download the Arctic ice data using the `get_icecurves()` function from the {hockeystick} package.

```{r}
#| echo: true
library(hockeystick)

arcticice <- get_icecurves()
```

##

Look at the data using the `head()` function.

```{r}
#| echo: true
arcticice |>
  head()
 
```
Arctic ice extent in millions of square kilometers every month since 1979.

## Make a time-series plot

```{r}
#| echo: true
arcticice |>
  plot_icecurves()
```

## Paleo ice core

```{r}
#| echo: true
library(hockeystick)

vostok <- get_paleo()

vostok |>
  head()

vostok |>
  tail()
```

##

```{r}
#| echo: true
vostok |>
  plot_paleo()
```

## All together

```{r}
#| echo: true
#| output-location: slide 
library(hockeystick)

climate_grid()
```
