---
title: "Climate swings"
format:
  revealjs:
    slide-number: c/t
    code-fold: false
    chalkboard: true
    theme: white
    footer: "GEO4251 Climate Change & Storms"
editor: source
---

##

-   The climate is constantly changing being pushed from outside as well as from inside
-   Orbital changes, volcanoes, solar activity
-   El Niño and other internal oscillations

##

-   Centuries to millennium
    - Orbital changes (Milankovitch cycles)
    - Shape of Earth’s orbit: eccentricity (100K)
    - Angle at which Earth’s axis is tilted with respect to Earth’s orbital plane: obliquity (41K)
    - Direction Earth’s axis of rotation is pointed: precession (26K)
    
##

-   Months to years
    - Large volcano eruptions: first cool, then warm

{{< video https://www.youtube.com/embed/7zoycrMusK4 >}} 2 minutes

  - Superposed epoch method

## {.smaller}

-   Changes to the sun
-   Times of maximum sunspot activity are times with more energy output from the sun (mostly ultraviolet) 

{{< video https://www.youtube.com/embed/wWIh3PvQscQ >}} 5 minutes

-   Solar changes and hurricanes in the Caribbean <https://myweb.fsu.edu/jelsner/PDF/Research/ElsnerJagger2008.pdf>

## {.smaller}

-   Daily sunspot counts from <https://www.sidc.be/silso/datafiles#total>
-   Data are in the file `Sunspots.txt` in your directory `data`
-   Get the values into your R session environment as a data frame called `ssn.df` using the `read_table()` function from the {readr} package

```{r, message=FALSE}
#| echo: true
ssn.df <- readr::read_table(file = here::here("data", "Sunspots.txt"),
                            col_names = c("Year", "Month", "Day", "YearF", 
                                          "SSN", "SSN_error", "Nobs", "X")) |>
  dplyr::filter(Year >= 1950, Year <= 2021)
```

##

-   Create a time series plot of daily sunspot numbers

```{r}
#| echo: true
ssn.df <- ssn.df |>
  dplyr::mutate(Date = as.Date(paste0(Year, "-", Month, "-", Day)))

library(ggplot2)

ggplot(data = ssn.df,
       mapping = aes(x = Date, y = SSN)) +
  geom_line() +
  geom_smooth(span = .05, method = "loess")
```


##

-   El Niño is part of an internal climate swing 

{{< video https://www.youtube.com/embed/WPA-KpldDVc >}} 4 minutes

##

-   The Ocean Niño Index (ONI) tracks the development and intensity of El Niño using ocean temperatures across the east tropical Pacific
-   It is the three-month average sea-surface temperature anomaly (relative to 1961-1990) in degrees Celsius for an area in the east tropical Pacific
-   Large positive values of the ONI indicate the occurrence of El Niño, large negative values indicate the occurrence of La Niña

##

-   Monthly values of the ONI are in your file `ONI.csv` in your directory `data`
-   Get the values into your R session environment as a data frame called `oni.df` using the `read_csv()` function from the {readr} package

```{r, message=FALSE}
#| echo: true
oni.df <- readr::read_csv(here::here("data", "ONI.csv")) 
oni.df
```

##

-   Add a date column called `Date` in `oni.df` using the `as.Date()` function combining the `Year` and `Month` columns

```{r}
#| echo: true
oni.df <- oni.df |>
  dplyr::mutate(Date = as.Date(paste0(Year, "-", Month, "-01")))

oni.df
```

## 

-   Make a time-series plot. Add lines corresponding to values greater than 2 and less than -1.5

```{r, message=FALSE}
#| echo: true
#| output-location: slide

ggplot(data = oni.df,
       mapping = aes(x = Date, y = ONI_C)) +
  geom_line() +
  geom_hline(yintercept = c(2, -1.5), col = "red") +
  labs(x = "Year", y = "ONI (°C)") +
  theme_minimal()
```

-   Values above 1°C indicate moderate El Niño conditions and values above 2°C indicate strong El Niño conditions
-   Note the upward trend toward even stronger El Niño conditions

##

-   Plot the upward trend in the strong El Niño conditions with the `geom_quantiles()` function setting `quantiles = .99`

```{r, message=FALSE}
#| echo: true
ggplot(data = oni.df,
       mapping = aes(x = Date, y = ONI_C)) +
  geom_line() +
  geom_quantile(quantiles = .99) +
  labs(x = "Year", y = "ONI (°C)") +
  theme_minimal()

```


## {.smaller}

-   The Southern Oscillation Index (SOI) tracks the development and intensity of El Niño using air pressure
-   El Niño - Southern Oscillation (ENSO)
-   The SOI is calculated from the monthly fluctuations in the air pressure between the island of Tahiti and the city of Darwin in Australia
-   Sustained negative values of the SOI indicate El Niño episodes

```{r}
r <- terra::rast()
terra::values(r) <- 0
tmap::tmap_mode("view")
tmap::tm_shape(r) + tmap::tm_raster(alpha = .1)
```

## {.smaller}

-   The SOI is a *standardized anomaly* of the mean sea level pressure difference. The averaging (base) period is the 60 years (1933--1992)

-   For each month $j = 1,2, \ldots, 12$ and each year $i = 1, \ldots, N$ let

$$
\Delta_{i,j} = P_{Tahiti} - P_{Darwin}
$$

then the SOI for year $i$ and month $j$ is

$$
\hbox{SOI}_{i,j} = 10\times\frac{\Delta_{i,j} - \bar{\Delta_j}}{\hbox{sd}(\Delta_j)}
$$

where $\bar{\Delta_j}$ is the 60-year average pressure difference for month $j$ and sd($\Delta_j$) is the 60-year standard deviation of the pressure difference for month $j$

## 

-   Data source: <http://www.bom.gov.au/climate/enso/soi_monthly.txt>
-   Get the data into your R session and create a `Date` column

```{r}
#| echo: true
soi.df <- readr::read_csv(file = here::here("data", "SOI.txt")) |>
  dplyr::mutate(Year = round(YearMo/100),
                Month = YearMo - Year * 100,
                Date = as.Date(paste0(Year, "-", Month, "-01")))
```

-   Make a time-series plot

```{r}
#| echo: true
#| output-location: slide
ggplot(data = soi.df,
       mapping = aes(x = Date, y = SOI_times10sd / 10)) +
  geom_line() +
  labs(x = "Year", y = "SOI (s.d.)") +
  theme_minimal()
```

## 

-   Use the `dplyr::left_join()` function to create a new data frame containing both the ONI and SOI for the months that have both values

```{r}
#| echo: true
df <- oni.df |>
  dplyr::left_join(soi.df) |>
  dplyr::select(Date, ONI_C, SOI_times10sd) |>
  na.omit()

df |>
  head()
```

## 

-   Based on what you know about how El Niño works what do you think the correlation between the ONI and SOI will be?
-   Compute the correlation between the ONI and SOI

```{r}
#| echo: true
cor(df$ONI_C, df$SOI_times10sd)
```

-   Make a plot

```{r}
#| echo: true
#| output-location: slide
ggplot(data = df,
       mapping = aes(x = ONI_C, 
                     y = SOI_times10sd / 10)) +
  geom_point() +
  geom_smooth(method = lm) +
  labs(x = "ONI (°C)", y = "SOI (s.d.)") +
  theme_minimal()
```

## 

-   More information about El Niño

<https://www.metoffice.gov.uk/research/climate/seasonal-to-decadal/gpc-outlooks/el-nino-la-nina/enso-description>

##

-   Tropical North Atlantic SST <https://psl.noaa.gov/data/climateindices/list/>
-   In the file `TNA_SST.txt` in directory `data`
-   Get the data into your R session and make a plot

```{r}
#| echo: true

TNA.df <- readr::read_table(file = here::here("data", "TNA_SST.txt"))

ggplot(data = TNA.df,
       mapping = aes(x = Year, y = Aug)) +
  geom_line() + 
  geom_smooth() +
  labs(y = "°C")
```


## 

-   North Atlantic Oscillation

{{< video https://www.youtube.com/embed/KOYJG7j4Iy8 >}} 3 minutes

-   Pacific Decadal Oscillation

{{< video https://www.youtube.com/embed/Sc3tOEcM0YE >}} 3 minutes
