---
title: "El Niño"
format:
  revealjs:
    slide-number: c/t
    code-fold: false
    chalkboard: true
bibliography: References.bib
editor: source
---

## 

How El Niño works

{{< video https://www.youtube.com/embed/WPA-KpldDVc >}}


## 

The Ocean Niño Index (ONI) gives and indication of the development and intensity of El Niño using temperatures at the ocean's surface across the east tropical Pacific.

Specifically the ONI is a three-month mean sea-surface temperature anomalies (relative to 1961-1990) in degrees Celsius for an area in the east tropical Pacific.

## 

The values are in the file `ONI.csv` in the directory `data`.

Get the data into your R session using the `read_csv()` function from the {readr} package.

```{r}
#| echo: true
oni.df <- readr::read_csv(here::here("data", "ONI.csv")) |>
  data.frame()
```

Add a date column using the `as.Date()` function.

```{r}
#| echo: true
oni.df <- oni.df |>
  dplyr::mutate(Date = as.Date(paste0(Year, "-", Month, "-01")))
```

## 

Make a time-series plot.

```{r}
#| echo: true
#| output-location: slide
library(ggplot2)

ggplot(data = oni.df,
       mapping = aes(x = Date, 
                     y = ONI_C)) +
  geom_line() +
  geom_hline(yintercept = c(2, -1.5), col = "red") +
  labs(x = "Year", y = "ONI (°C)") +
  theme_minimal()
```

## 

The Southern Oscillation Index (SOI) gives an indication of the development and intensity of El Niño and La Niña events using air pressure.

The SOI is calculated from the monthly fluctuations in the air pressure between Tahiti and Darwin, Australia.

Sustained negative values of the SOI indicate El Niño episodes.

## 

The SOI is a *standardized anomaly* of the mean sea level pressure difference. The averaging (base) period is the 60 years (1933--1992).

For each month $j = 1,2, \ldots, 12$ and each year $i = 1, \ldots, N$ let $\Delta_{i,j}$ = $P_{Tahiti} - P_{Darwin}$, then the SOI for year $i$ and month $j$ is

$$
\hbox{SOI}_{i,j} = 10\times\frac{\Delta_{i,j} - \bar{\Delta_j}}{\hbox{sd}(\Delta_j)}
$$

where $\bar{\Delta_j}$ is the 60-year average pressure difference for month $j$ and sd($\Delta_j$) is the 60-year standard deviation of the pressure difference for month $j$.

## 

Data source: <http://www.bom.gov.au/climate/enso/soi_monthly.txt>

Get the data into your R session.

```{r}
#| echo: true
soi.df <- readr::read_csv(file = here::here("data", "SOI.txt")) |>
  dplyr::mutate(Year = round(YearMo/100),
                Month = YearMo - Year * 100,
                Date = as.Date(paste0(Year, "-", Month, "-01")))
```

Make a time-series plot.

```{r}
#| echo: true
#| output-location: slide
ggplot(data = soi.df,
       mapping = aes(x = Date, y = SOI_times10sd / 10)) +
  geom_line() +
  labs(x = "Year", y = "SOI (s.d.)") +
  theme_minimal()
```

## 

Use the `left_join()` function to create a new data frame containing both the ONI and SOI for the months that have both values.

```{r}
#| echo: true
df <- oni.df |>
  dplyr::left_join(soi.df) |>
  dplyr::select(Date, ONI_C, SOI_times10sd) |>
  na.omit()

head(df)
```

## 

Based on what you know about how El Niño works what do you think the correlation between the ONI and SOI will be?

Compute the correlation between the ONI and SOI.

```{r}
#| echo: true
cor(df$ONI_C, df$SOI_times10sd)
```

Make a plot.

```{r}
#| echo: true
#| output-location: slide
ggplot(data = df,
       mapping = aes(x = ONI_C, 
                     y = SOI_times10sd / 10)) +
  geom_point() +
  geom_smooth(method = lm) +
  labs(x = "ONI (C)", y = "SOI (s.d.)") +
  theme_minimal()
```

## 

More information <https://www.metoffice.gov.uk/research/climate/seasonal-to-decadal/gpc-outlooks/el-nino-la-nina/enso-description>
