---
title: "Lab 4"
author: "Your Name"
format: pdf
editor: source
---

## 1 
Write code to create a time series graph showing the the number of Florida hurricanes each year
```{r}
CONUS.df <- readr::read_table(file = here::here("data", "CONUS_Counts.txt"),
                              show_col_types = FALSE)
library(ggplot2)

ggplot(data = CONUS.df,
       mapping = aes(x = Year, y = FL)) +
  geom_col()
```

## 2 
Write code to compute the mean, variance, and ratio of the variance to the mean for the annual number of Florida hurricanes
```{r}
CONUS.df |>
  dplyr::summarize(NYears = dplyr::n(),
                   Avg = mean(FL),
                   Var = var(FL),
                   Ratio = Var/Avg)
```

## 3 
Write code to determine the total number of hurricanes in each of the three regions (Gulf coast, Florida, or East coast) since 1851 
```{r}
CONUS.df |>
  dplyr::summarize(nH_G = sum(G),
                   nH_FL = sum(FL),
                   nH_E = sum(E))
```

## 4
Write code to compute the correlation between the annual count of Gulf coast hurricanes and the annual count of East coast hurricanes
```{r}
CONUS.df |>
  dplyr::summarize(r = cor(G, E))
```
