---
title: "Greenhouse gas emissions"
format: 
  revealjs:
    slide-number: c/t
    code-fold: false
    chalkboard: true
bibliography: References.bib
editor: source
---


## Get the emissions data

Download the emissions data using the `get_emissions()` function from the {hockeystick} package.

```{r}
#| echo: true
library(hockeystick)

emits <- hockeystick::get_emissions(write_cache = TRUE)
```

##

Look at the first six rows of the data frame using the `head()` function.

```{r}
#| echo: true
emits |>
  head()
```

## Make a time-series plot

```{r}
#| echo: true
#| fig-cap: "Annual total emissions worldwide"
hockeystick::plot_emissions(emits)
```

## By country/region in 2020
```{r}
#| echo: true
emits |>
  dplyr::filter(year == 2020) |>
  dplyr::select(country, co2, co2_per_capita) |>
  dplyr::arrange(desc(co2))
```

China, United States, India, Russia, Japan, Iran, Germany, Saudi Arabia, South Korea, Indonesia, Canada, Brazil, South Africa, Turkey, Australia, United Kingdom, Italy, Poland, Kazakhstan, France, Taiwan, Malaysia