---
title: "El Niño"
format:
  revealjs:
    slide-number: c/t
    code-fold: false
    chalkboard: true
bibliography: References.bib
editor: source
---


## 

How El Niño works

{{< video https://www.youtube.com/embed/WPA-KpldDVc >}}

<https://youtu.be/WPA-KpldDVc>

##

Three-month mean sea-surface temperature anomalies (relative to 1961-1990) in degrees Celsius for an area in the east tropical Pacific. The values are in the file `enso_ONI.csv` in the directory `data`.

Get the data into your R session using the `read_csv()` function from the {readr} package.

```{r}
#| echo: true
df <- readr::read_csv(here::here("data", "enso_ONI.csv")) |>
  data.frame()
```
Add a date column using the `as.Date()` function.

```{r}
#| echo: true
df <- df |>
  dplyr::mutate(Date = as.Date(paste0(Year, "-", Month, "-01")))
```

##

Make a time-series plot.

```{r}
#| echo: true
#| output-location: slide
library(ggplot2)

ggplot(data = df,
       mapping = aes(x = Date, 
                     y = ONI_C)) +
  geom_line() +
  geom_hline(yintercept = c(2, -1.5), col = "red") +
  labs(x = "Year", y = "ONI (°C)") +
  theme_minimal()
```

##

The Southern Oscillation Index (SOI) gives an indication of the development and intensity of El Niño or La Niña events in the Pacific Ocean. The SOI is calculated from the monthly fluctuations in the air pressure between Tahiti and Darwin, Australia.

Sustained negative values of the SOI indicate El Niño episodes. These negative values are accompanied by sustained warming of the central and eastern tropical Pacific Ocean.

##

The SOI is a standardized anomaly of the mean sea level pressure difference. The base period is 60 years (1933–1992).

Let $\Delta$ = $P_{Tahiti} - P_{Darwin}$, then for each month $j = 1,2, \ldots, 12$

$$
\hbox{SOI}_j = 10\times\frac{\Delta_j - \bar{\Delta_j}}{\hbox{sd}(\Delta_j)}
$$

where $\bar{\Delta_j}$ is the 60-year average pressure difference for month $j$.

Data source: <http://www.bom.gov.au/climate/enso/soi_monthly.txt>

Get the data into your R session.

```{r}
#| echo: true
soi.df <- readr::read_csv(file = here::here("data", "SOI.txt")) |>
  dplyr::mutate(Year = round(YearMo/100),
                Month = YearMo - Year * 100,
                Date = as.Date(paste0(Year, "-", Month, "-01")))
```

##

Make a time-series plot.

```{r}
#| echo: true
ggplot(data = soi.df,
       mapping = aes(x = Date, y = SOI_times10sd / 10)) +
  geom_line() +
  labs(x = "Year", y = "SOI (s.d.)") +
  theme_minimal()
```

## 

Use left_join on Date and then compute correlation between SOI and ONI after removing NA.

##

More information
<https://www.metoffice.gov.uk/research/climate/seasonal-to-decadal/gpc-outlooks/el-nino-la-nina/enso-description>
