Tornadoes
=========

This week we turn our attention to tornadoes. 

Tornadoes occur seemly random across the country; so a first step toward understanding them is to make a map showing where they occur.

There are many ways to make a map with R. One easy way is with the functions in the **ggmap** package. 

Let's begin by making a map of the campus.

The first line of code makes the functions from **ggmap** available. The second line uses the geocode() function to locate the Bellamy Building. 
```{r}
library("ggmap")
loc = geocode("Bellamy Building Florida State University")
loc
```

Next we use the get_map() function to call the google map server and center the map on loc. The map is saved in the object called Map.
```{r}
Map = get_map(location = loc, 
              source = "google",
              zoom = 15,
              color = "bw",
              maptype = "terrain")
```

Finally we use the ggmap() function to render the map to the display.
```{r}
ggmap(Map, dev = "extent") +
    labs(x = expression(paste("Longitude (", degree, "W)")), 
       y = expression(paste("Latitude (", degree, "N)")))
```

Easy!

Try making a map of Tallahassee or your home town.
