My Term Project
===============
author: My Name

My term project is worth 30% of my semester grade. It will be graded on a 100 point scale.

I rename the file replacing "MyName" in the MyName_TermProject.Rmd file name to my name. **This is worth 5 points.**

Next to the word author: change My Name to my name. **This is worth 5 points.**

Note: Select the Packages tab (lower right) > Install. Type: ggplot2, then click install to install the needed package.

Basic idea behind R markdown (Rmd): code + narrative = scientific report (computer language [R] + authoring language [MD])

### 1. Introduction

Here is where I type a description of my dataset. **This is worth 10 points.**

### 2. Get the Data

Here is where I show the code used to get the data in R. **This is worth 20 points.**

```{r}
L = 'http://myweb.fsu.edu/jelsner/data/FLMonthlyT.txt'
df = read.table(L, header = TRUE)
head(df)
```

### 3. Graph the Data

Here is where I show the code to make a graph of the data. I will make sure I label the axes appropriately. **This is worth 20 points.**

```{r}
library(ggplot2)
ggplot(df, aes(x = Year, y = Annual)) +
  geom_line() + 
  ylab("Temperature Anomaly (C)")
```

### 4. Draw a Trend Line

Here is where I show the code that adds a trend line to the graph. **This is worth 20 points.**

```{r}
ggplot(df, aes(x = Year, y = Annual)) +
  geom_line() + geom_smooth() +
  ylab("Temperature Anomaly (C)")
```

### 5. Describe the Graph

Here is where I write a few sentences describing the graph and trend line. **This is worth 10 points.**

### 6. Generate an HTML file and send it to jelsner@fsu.edu

Here is where I select the Knit HTML button above. I then open my email client and send the HTML file to Professor Elsner. 

**Note: If I can't produce an HTML file, I send the Rmd file instead.**
