### Is Tallahassee Getting Hotter?

Begin by getting your dataset into R. The first line of code sets the location of the data as a string of characters (letters, slashes, etc). The second line of code inputs the data as a data frame. The argument header = TRUE lets R know that the dataset has a top row that identifies the columns by name (header record).
```{r}
L = 'http://myweb.fsu.edu/jelsner/data/TLHT.txt'
df = read.table(L, header = TRUE)
head(df)
```

Next load the packages that contain functions that make it easy to manipulate the data and to make a graph.
```{r, message=FALSE}
library(ggplot2)
library(dplyr)
library(magrittr)
```

Arrange the data. To make a new data frame (Annual) take the data frame (df) then (%>%) group by year then average the daily high temperatures the plot.

```{r}
Annual = df %>% 
  group_by(Year) %>%
  summarize(Avg = mean(Tmax))
```

Make the graph. Use the ggplot()on to graph the data in the Annual data frame.
```{r}
ggplot(Annual, aes(x = Year, y = Avg)) +
  geom_point() + 
  geom_line() +
  geom_smooth(method = lm, size = 3)
```

Quantify the trend.
```{r}
coef(lm(Avg ~ Year, data = Annual))[2]
```

[More Heat](http://www.wunderground.com/climate/videos.asp?video=extreme_weather_101_heat)

