### Is Tallahassee Getting Hotter?

First, get your data into R. The first line of code sets the location of the data as a string of characters (letters, slashes, etc). The second line of code inputs the data as a data frame. The argument header = TRUE lets R know that the dataset has a top row that identifies the columns by name (header record).
```{r}
L = 'http://myweb.fsu.edu/jelsner/data/TLHT.txt'
df = read.table(L, header = TRUE)
```

Check by selecting the df object under the Environment tab.

Next load the functions that make it easy to manipulate the data and to make a graph.
```{r, message=FALSE}
library(ggplot2)
library(dplyr)
```

Arrange the data. To make a new data frame (Annual) take the data frame (df) then (%>%) group by year then average the daily high temperatures.
```{r}
Annual = df %>% 
  group_by(Year) %>%
  summarize(Avg = mean(Tmax))
```

Make the graph. Use the ggplot() to graph the data in the Annual data frame.
```{r}
ggplot(Annual, aes(x = Year, y = Avg)) +
  geom_point() + 
  geom_line() +
  geom_smooth(method = lm, size = 3)
```

Quantify the trend.
```{r}
coef(lm(Avg ~ Year, data = Annual))[2]
```

### Is the Frequency of Extremely Hot Days Increasing?

Here we define a hot day as one in which the high temperatures reaches 100F.

```{r}
df %>%
  group_by(Year) %>%
  summarise(N100 = sum(Tmax >= 100)) %>%
ggplot(., aes(x = Year, y = N100)) +
  geom_bar(stat = "identity") +
  ylab("Number of days in Tallahassee at or exceeding 100F")
```

Make it prettier.
```{r}
df %>%
  group_by(Year) %>%
  summarise(N100 = sum(Tmax >= 100)) %>%  
ggplot(., aes(x = Year, y = N100, fill = N100)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  scale_fill_continuous(low = 'orange', high = 'red') +
  geom_text(aes(label = N100), vjust = 1.5, size = 3) +
  ylab("Number of days in Tallahassee at or exceeding 100F") +
  scale_x_continuous(breaks = seq(1950, 2013, 10)) +
  xlab("") +
  theme(axis.text.x  = element_text(size = 11), 
        legend.position = "none")
```

The hots are getting hotter.

[More Heat](http://www.wunderground.com/climate/videos.asp?video=extreme_weather_101_heat)

